---
title: Assignment 1
subtitle: Smoothing
description:
  Implement and test kernel density estimation and bivariate smoothing spline
  with bandwidth selection, comparing results to R's built-in functions.
---

The first assignment topic is smoothing. If you draw the topic "smoothing" at
the oral exam, you will have to present a solution of one of the two assignments
below.

For all assignments in the course you should have the following five points in
mind:

- How can you test that your implementation is correct?
- Can you implement alternative solutions?
- Can the code be restructured e.g. by modularization, abstraction or object
  oriented programming to improve generality, extensibility, and readability?
- How does the implementation perform (benchmarking)?
- Where are the bottlenecks (profiling), and what can you do about them?

Some of these points are not covered yet (or you just don't master them right
now), and you may need to return to your solution later in the course to improve
and refine it. Moreover, the five points are not independent. Alternative
solutions may provide ways of testing the implementation. Restructuring of code
may improve or hurt performance. Optimization of performance may improve or hurt
readability. And so on.

# A: Density Estimation

Implement a kernel density estimator using the Epanechnikov kernel,
$$
  K(x) = \frac{3}{4}(1 - x^2) 1_{[-1, 1]}(x),
$$
and implement two bandwidth-selection methods: an AMISE plug-in method and a
cross-validation method. Test your implementation using both real and simulated
data, and investigate how the two methods behave. Compare your results with
`density()` and suitable bandwidth selectors provided by R. Make your
implementation as general as possible.

::: {.callout-tip}
When comparing your implementation with R's built-in `density()`{.R}, make sure
to read the documentation carefully, in particular the documentation of the `bw`
argument.
:::

## Approach and course references

The implementation follows the vectorized estimator in the course's
[`R/kernel.R`](https://github.com/math-ku/compstat/blob/main/R/kernel.R).
[Lecture 2](https://github.com/math-ku/compstat/blob/main/slides/lecture2.qmd)
provides the AMISE expression and discusses normal-reference bandwidths and
Sheather--Jones selection. The [Week 2 solutions, Exercises 2.1--2.5](https://math-ku.github.io/compstat/exercises/U2.html)
provide the Epanechnikov constants, the `log(F12)` example, bandwidth scaling,
and benchmarking exercises. The bandwidth selectors below are implemented
here from their mathematical criteria; they are not calls to R's selectors.

All computations use base R. The real data are already available locally in
`Data/infrared.txt`, so rendering does not require a download.

## Kernel and density estimator

For observations $X_1,\ldots,X_n$, define
$$
\widehat f_h(t)=\frac{1}{nh}\sum_{i=1}^n K\left(\frac{t-X_i}{h}\right).
$$
Here $h$ is the **support half-width** of each kernel. The Epanechnikov
kernel is nonnegative, symmetric, and has
$$
\int K(u)\,du=1,\qquad
\mu_2(K)=\int u^2K(u)\,du=\frac15,\qquad
R(K)=\int K(u)^2\,du=\frac35.
$$
Thus $\widehat f_h$ is itself a density. Small $h$ produces narrow peaks and
large sampling variation; large $h$ smooths more but can obscure structure.

The function accepts any finite numeric sample and either a custom evaluation
grid or a grid size. Missing values are removed only when requested. Kernel
evaluation and bandwidth selection are separate functions, so the estimator
can also accept another vectorized density kernel. For a kernel of unbounded
support, supply a suitably wide grid explicitly.

```{r}
#| label: kde-functions
clean_sample <- function(x, na.rm = FALSE, min_n = 1L) {
  if (!is.numeric(x) || !is.null(dim(x)))
    stop("x must be a numeric vector.")
  if (!is.logical(na.rm) || length(na.rm) != 1L || is.na(na.rm))
    stop("na.rm must be TRUE or FALSE.")
  if (na.rm) x <- x[!is.na(x)]
  if (length(x) < min_n || any(!is.finite(x)))
    stop("Too few observations, or missing/infinite values in x.")
  x
}

positive_scalar <- function(x, name) {
  if (!is.numeric(x) || length(x) != 1L || !is.finite(x) || x <= 0)
    stop(name, " must be a finite positive number.")
}

epanechnikov <- function(u) {
  # Subsetting avoids evaluating a polynomial at very large distances.
  ans <- numeric(length(u))
  inside <- abs(u) <= 1
  ans[inside] <- 0.75 * (1 - u[inside]^2)
  ans
}

kde <- function(x, h, grid = NULL, m = 512L,
                kernel = epanechnikov, na.rm = FALSE) {
  x <- clean_sample(x, na.rm)
  positive_scalar(h, "h")
  if (is.null(grid)) {
    positive_scalar(m, "m")
    if (m < 2 || m != floor(m)) stop("m must be an integer >= 2.")
    grid <- seq(min(x) - h, max(x) + h, length.out = m)
  }
  if (!is.numeric(grid) || length(grid) == 0L || any(!is.finite(grid)))
    stop("grid must contain finite numeric evaluation points.")
  y <- vapply(grid, function(t) mean(kernel((t - x) / h)) / h,
              numeric(1))
  list(x = grid, y = y, h = h, n = length(x))
}

# Trapezoidal integration on an increasing, possibly nonuniform grid.
trapz <- function(x, y) {
  stopifnot(length(x) == length(y), length(x) >= 2L, all(diff(x) > 0))
  sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2)
}
```

## AMISE plug-in bandwidth

Write $R(f'')=\int f''(t)^2\,dt$. Substituting the kernel constants into the
lecture's bias--variance approximation gives
$$
\operatorname{AMISE}(h)=\frac{3}{5nh}+\frac{h^4}{100}R(f''),
\qquad
h_{\mathrm{AMISE}}=\left(\frac{15}{nR(f'')}\right)^{1/5}.
$$
Differentiating with respect to $h$ gives the minimizer. The unknown quantity
is the curvature $R(f'')$: a density with narrow peaks needs a smaller bandwidth.

We implement a **normal-reference plug-in method**. For a normal reference
density with standard deviation $s$, $R(f'')=3/(8\sqrt\pi s^5)$, hence
$$
\widehat h_{\mathrm{NR}}=(40\sqrt\pi)^{1/5}\widehat s\,n^{-1/5}.
$$
By default, $\widehat s=\min\{\operatorname{sd}(X),\operatorname{IQR}(X)/1.34\}$,
using the robust scale convention in Lecture 2; if the IQR is zero we use the
standard deviation. Setting `robust = FALSE` uses the standard deviation alone.
This is a simple parametric plug-in estimate of curvature, not the more elaborate
Sheather--Jones pilot-derivative method. It can oversmooth multimodal data because
a single normal reference does not describe their local curvature. Constant
samples have no positive estimated scale, so automatic selection raises an error;
the KDE itself still works with a user-supplied positive bandwidth.

```{r}
#| label: amise-selector
bw_amise <- function(x, robust = TRUE, na.rm = FALSE) {
  x <- clean_sample(x, na.rm, min_n = 2L)
  s <- sd(x)
  if (robust && IQR(x) > 0) s <- min(s, IQR(x) / 1.34)
  positive_scalar(s, "Estimated scale")
  (40 * sqrt(pi))^(1 / 5) * s * length(x)^(-1 / 5)
}
```

## Leave-one-out least-squares cross-validation

Expanding the integrated squared error gives
$$
\operatorname{ISE}(h)=\int\widehat f_h^2
 -2\int\widehat f_h f+\int f^2.
$$
The last term does not depend on $h$. Estimate the middle integral by predicting
each observation from all the **other** observations:
$$
\operatorname{CV}(h)=\int\widehat f_h(t)^2\,dt
 -\frac{2}{n}\sum_{i=1}^n\widehat f_{h,-i}(X_i),\qquad
\widehat f_{h,-i}(X_i)=\frac{1}{(n-1)h}\sum_{j\ne i}K\left(\frac{X_i-X_j}{h}\right).
$$
For independent observations, its expectation equals MISE minus $\int f^2$.
Including the observation itself in the second term would reward narrow spikes
at the data points and is therefore incorrect. Negative CV scores are possible:
the constant $\int f^2$ has been omitted.

For this symmetric kernel let $C(v)=\int K(u)K(u+v)\,du$. Direct integration
of the two quadratic polynomials over their overlapping supports gives, with
$a=|v|$,
$$
C(v)=\frac{3}{160}(2-a)^3(a^2+6a+4)\,1_{\{a\le2\}}.
$$
Consequently we can evaluate the first integral exactly, without an integration
grid. Storing just the distances $d_{ij}=|X_i-X_j|$ for $i<j$ gives
$$
\operatorname{CV}(h)=
\frac{nC(0)+2\sum_{i<j}C(d_{ij}/h)}{n^2h}
-\frac{4\sum_{i<j}K(d_{ij}/h)}{n(n-1)h}.
$$
The diagonal belongs in the first term only. Repeated observations remain
separate observations, even when their pairwise distance is zero.

```{r}
#| label: cv-selector
epan_convolution <- function(v) {
  a <- abs(v)
  ans <- numeric(length(a))
  inside <- a <= 2
  z <- a[inside]
  ans[inside] <- 3 / 160 * (2 - z)^3 * (z^2 + 6 * z + 4)
  ans
}

make_cv_score <- function(x, na.rm = FALSE) {
  x <- clean_sample(x, na.rm, min_n = 2L)
  n <- length(x)
  distances <- as.vector(dist(x))
  function(h) {
    positive_scalar(h, "h")
    u <- distances / h
    (n * 3 / 5 + 2 * sum(epan_convolution(u))) / (n^2 * h) -
      4 * sum(epanechnikov(u)) / (n * (n - 1) * h)
  }
}

bw_cv <- function(x, interval = NULL, n_grid = 100L, na.rm = FALSE) {
  x <- clean_sample(x, na.rm, min_n = 2L)
  if (is.null(interval)) interval <- bw_amise(x) * c(0.05, 4)
  if (!is.numeric(interval) || length(interval) != 2L ||
      any(!is.finite(interval)) || interval[1] <= 0 ||
      interval[2] <= interval[1]) stop("Invalid positive search interval.")
  positive_scalar(n_grid, "n_grid")
  if (n_grid < 3 || n_grid != floor(n_grid))
    stop("n_grid must be an integer >= 3.")
  score <- make_cv_score(x)
  log_h <- seq(log(interval[1]), log(interval[2]), length.out = n_grid)
  values <- vapply(exp(log_h), score, numeric(1))
  # CV need not be unimodal: refine every local minimum detected on the grid.
  middle <- seq.int(2L, n_grid - 1L)
  local <- middle[values[middle] <= values[middle - 1L] &
                    values[middle] <= values[middle + 1L]]
  refined <- vapply(local, function(i) {
    optimize(function(z) score(exp(z)), log_h[c(i - 1L, i + 1L)],
             tol = 1e-6)$minimum
  }, numeric(1))
  candidates <- exp(c(log_h, refined))
  scores <- c(values, vapply(exp(refined), score, numeric(1)))
  best <- which.min(scores)
  h <- candidates[best]
  boundary <- h <= interval[1] * (1 + 1e-5) ||
              h >= interval[2] * (1 - 1e-5)
  if (boundary) warning("CV minimum is at a search boundary; inspect/expand interval.")
  list(h = h, score = scores[best], boundary = boundary,
       curve = data.frame(h = exp(log_h), CV = values), interval = interval)
}
```

The search is over a broad interval relative to the plug-in bandwidth, on a log
scale so small bandwidths are resolved. Grid search followed by local refinement
reduces the risk of choosing the wrong local minimum, but does not prove global
optimality. The returned curve and boundary flag make the search inspectable;
`interval` and `n_grid` can be changed. With many ties, CV can favor bandwidths
approaching zero, so a boundary result should not be interpreted as a reliable
positive optimum. The method assumes continuous, independent observations.

## Correctness checks and comparison with `density()`

R's `?density` defines `bw` as the standard deviation of the smoothing kernel.
Our scaled kernel has variance $h^2/5$, so an **identical Epanechnikov estimate**
requires `density(x, bw = h / sqrt(5), kernel = "epanechnikov")`.
The two methods should be close, rather than bitwise equal, because R bins the
data and uses an FFT approximation. This is the scaling issue in Exercise 2.3.

We test kernel moments, normalization, a direct double-loop implementation,
translation/scale equivariance, input handling, and the CV formula against
numerical integration plus explicit leave-one-out refits. The latter check is
independent of the convolution implementation.

```{r}
#| label: correctness-checks
moments <- c(
  mass = integrate(epanechnikov, -1, 1)$value,
  mean = integrate(function(u) u * epanechnikov(u), -1, 1)$value,
  second = integrate(function(u) u^2 * epanechnikov(u), -1, 1)$value,
  squared = integrate(function(u) epanechnikov(u)^2, -1, 1)$value
)
stopifnot(max(abs(moments - c(1, 0, 1 / 5, 3 / 5))) < 1e-10)

x_test <- c(-1.3, -0.2, -0.2, 0.7, 1.6)
h_test <- 0.8
t_test <- seq(-2.5, 2.8, length.out = 501)
direct <- numeric(length(t_test))
for (i in seq_along(t_test)) {
  for (j in seq_along(x_test)) {
    u <- (t_test[i] - x_test[j]) / h_test
    if (abs(u) <= 1) direct[i] <- direct[i] + 0.75 * (1 - u^2)
  }
}
direct <- direct / (length(x_test) * h_test)
fit_test <- kde(x_test, h_test, grid = t_test)
stopifnot(max(abs(direct - fit_test$y)) < 1e-12,
          all(fit_test$y >= 0),
          max(abs(kde(3 + 2 * x_test, 2 * h_test,
                      grid = 3 + 2 * t_test)$y - fit_test$y / 2)) < 1e-12)
fine <- kde(x_test, h_test, m = 20001)
stopifnot(abs(trapz(fine$x, fine$y) - 1) < 1e-6)
stopifnot(isTRUE(all.equal(kde(c(x_test, NA), h_test, na.rm = TRUE),
                          kde(x_test, h_test))))
throws <- function(expr) inherits(tryCatch(expr, error = identity), "error")
stopifnot(throws(kde(x_test, 0)), throws(kde(c(1, NA), 1)),
          throws(kde(c(1, Inf), 1)), throws(bw_amise(rep(1, 5))),
          throws(bw_cv(1)))

# Integrate piecewise at all kernel boundaries to avoid missing narrow pieces.
numerical_cv <- function(x, h) {
  knots <- sort(unique(c(x - h, x + h)))
  square_integral <- sum(vapply(seq_len(length(knots) - 1L), function(i) {
    integrate(function(t) kde(x, h, grid = t)$y^2,
              knots[i], knots[i + 1L], rel.tol = 1e-9)$value
  }, numeric(1)))
  loo <- vapply(seq_along(x), function(i) {
    kde(x[-i], h, grid = x[i])$y
  }, numeric(1))
  square_integral - 2 * mean(loo)
}
cv_check <- vapply(c(0.15, 0.8, 2), function(h) {
  abs(make_cv_score(x_test)(h) - numerical_cv(x_test, h))
}, numeric(1))
stopifnot(max(cv_check) < 1e-7)
cat("All correctness checks passed.\n")
```

## Real data: infrared flux

Following Exercise 2.2, estimate the density of $Y=\log(F12)$. This is the
density on the log scale; a density on the original positive flux scale would
be $\widehat f_{F12}(z)=\widehat f_Y(\log z)/z$ for $z>0$.

```{r}
#| label: infrared-data
# Support both rendering from Assignments/ and running chunks from the project root.
data_paths <- c("../Data/infrared.txt", "Data/infrared.txt")
data_path <- data_paths[file.exists(data_paths)][1]
if (is.na(data_path)) stop("Cannot find Data/infrared.txt.")
infrared <- read.table(data_path, header = TRUE)
stopifnot(all(is.finite(infrared$F12)), all(infrared$F12 > 0))
x_real <- log(infrared$F12)
h_real <- bw_amise(x_real)
cv_real <- bw_cv(x_real)

# Compare the actual KDEs at identical bandwidths and identical grid points.
r_fit <- density(x_real, bw = h_real / sqrt(5), kernel = "epanechnikov",
                 from = min(x_real) - h_real, to = max(x_real) + h_real, n = 4096)
own_fit <- kde(x_real, h_real, grid = r_fit$x)
max_difference <- max(abs(own_fit$y - r_fit$y))
stopifnot(max_difference < 1e-3 * max(own_fit$y))
data.frame(n = length(x_real), distinct = length(unique(x_real)),
           h_AMISE = h_real, h_CV = cv_real$h,
           CV_boundary = cv_real$boundary, max_density_difference = max_difference)
```

### Comparing bandwidth selectors on a common scale

R's `bw.nrd()`, `bw.nrd0()`, `bw.SJ()`, and `bw.ucv()` are **Gaussian-kernel**
selectors (see `?bw.nrd`). To transfer their bandwidths $b_G$ to our kernel,
use the ratio of the AMISE constants:
$$
h_E=\left(\frac{R(K_E)/\mu_2(K_E)^2}
 {R(K_G)/\mu_2(K_G)^2}\right)^{1/5}b_G
=(30\sqrt\pi)^{1/5}b_G.
$$
For `bw.ucv()` we explicitly use the same search interval as our selector,
converted to Gaussian units, and tighten the numerical tolerance. R's default
interval produced boundary warnings in preliminary runs on these simulations.
R still uses its own binned Gaussian objective and optimization algorithm.

This conversion transfers an AMISE smoothing scale; it does not turn a Gaussian
CV optimizer into the exact Epanechnikov CV optimizer. Once $h_E$ is chosen,
`h_E / sqrt(5)` is the bandwidth to pass to R's Epanechnikov `density()`.

```{r}
#| label: selector-comparison
gaussian_to_epan <- (30 * sqrt(pi))^(1 / 5)
select_bandwidths <- function(x, cv = bw_cv(x)) {
  c(AMISE = bw_amise(x), CV = cv$h,
    R_nrd = gaussian_to_epan * bw.nrd(x),
    R_nrd0 = gaussian_to_epan * bw.nrd0(x),
    R_SJ = gaussian_to_epan * bw.SJ(x),
    R_ucv = gaussian_to_epan * bw.ucv(x,
      lower = cv$interval[1] / gaussian_to_epan,
      upper = cv$interval[2] / gaussian_to_epan,
      tol = 1e-6 * bw.nrd(x)))
}
real_bandwidths <- select_bandwidths(x_real, cv_real)
data.frame(method = names(real_bandwidths), h = unname(real_bandwidths),
           R_epanechnikov_bw = unname(real_bandwidths) / sqrt(5))

# The small difference from bw.nrd is its rounded Gaussian constant 1.06.
stopifnot(abs(h_real / real_bandwidths["R_nrd"] - 1) < 0.001)
```

```{r}
#| label: real-density-plots
#| fig-width: 9
#| fig-height: 4.5
old_par <- par(mfrow = c(1, 2))
cols <- c("#0072B2", "#D55E00", "#009E73", "#CC79A7", "#E69F00", "#555555")
grid_real <- seq(min(x_real) - max(real_bandwidths),
                 max(x_real) + max(real_bandwidths), length.out = 1024)
real_curves <- vapply(real_bandwidths, function(h) {
  kde(x_real, h, grid = grid_real)$y
}, numeric(length(grid_real)))
hist(x_real, probability = TRUE, breaks = "FD", col = "grey92", border = "white",
     main = "Infrared flux on log scale", xlab = "log(F12)",
     xlim = range(grid_real), ylim = c(0, max(real_curves) * 1.1))
matlines(grid_real, real_curves, col = cols, lty = seq_along(cols), lwd = 2)
legend("topright", names(real_bandwidths), col = cols, lty = seq_along(cols),
       lwd = 2, cex = 0.75, bty = "n")
plot(cv_real$curve, type = "l", log = "x", xlab = "Support bandwidth h",
     ylab = "Least-squares CV", main = "Inspecting the CV minimum")
abline(v = c(h_real, cv_real$h), col = cols[1:2], lty = 2)
legend("topright", c("AMISE", "CV"), col = cols[1:2], lty = 2, bg = "white")
par(old_par)
```

For this sample, the CV bandwidth is `r round(cv_real$h / h_real, 3)` times
the normal-reference bandwidth. The curves show which features persist across
smoothing choices. AMISE and the converted `bw.nrd()` nearly coincide because
they use the same reference scale. There is no known true flux density here,
so visual appearance alone cannot establish which method has the smallest error.
There are `r length(x_real)` observations but only
`r length(unique(x_real))` distinct log flux values. CV chooses about
`r round(cv_real$h, 3)` versus `r round(h_real, 3)` for AMISE, producing many more
small peaks. The CV curve is shallow over a substantial bandwidth range, and
repeated recorded values can encourage small bandwidths. Thus those additional
peaks should not automatically be interpreted as separate physical populations.

## Simulated data: accuracy and variability

Lecture 2 suggests Gaussian mixtures to investigate bandwidth selection. We use
both $N(0,1)$ and the separated mixture
$0.5N(-1.5,0.5^2)+0.5N(1.5,0.5^2)$. The normal case suits the reference assumption;
the mixture has much sharper local features than its overall spread suggests.

For each distribution and $n\in\{200,500\}$, repeat the experiment 20 times with
a fixed seed. Every method uses the same sample and the same Epanechnikov KDE.
Numerically integrated squared errors are computed on a grid covering both the
KDE supports and essentially all of the true density. Mean ISE estimates MISE;
the Monte Carlo standard error records the uncertainty from these 20 repetitions.
These are illustrative experiments, not precise universal rankings.

```{r}
#| label: simulation-study
set.seed(20260910)
simulations <- list(
  Normal = list(r = function(n) rnorm(n), d = dnorm),
  Bimodal = list(
    r = function(n) rnorm(n, mean = sample(c(-1.5, 1.5), n, replace = TRUE), sd = 0.5),
    d = function(t) 0.5 * dnorm(t, -1.5, 0.5) + 0.5 * dnorm(t, 1.5, 0.5)
  )
)
B <- 20L
rows <- list()
examples <- list()
for (case in names(simulations)) {
  model <- simulations[[case]]
  for (n in c(200L, 500L)) {
    for (b in seq_len(B)) {
      x <- model$r(n)
      cv <- bw_cv(x)
      hs <- select_bandwidths(x, cv)
      grid <- seq(min(-8, min(x) - max(hs)), max(8, max(x) + max(hs)),
                  length.out = 2048)
      truth <- model$d(grid)
      estimates <- vapply(hs, function(h) kde(x, h, grid = grid)$y,
                          numeric(length(grid)))
      ise <- apply(estimates, 2, function(y) trapz(grid, (y - truth)^2))
      rows[[length(rows) + 1L]] <- data.frame(
        distribution = case, n = n, repetition = b,
        method = names(hs), h = unname(hs), ISE = unname(ise),
        CV_boundary = cv$boundary
      )
      if (n == 500L && b == 1L)
        examples[[case]] <- list(grid = grid, truth = truth, estimates = estimates)
    }
  }
}
simulation_results <- do.call(rbind, rows)
groups <- split(simulation_results, interaction(simulation_results$distribution,
                 simulation_results$n, simulation_results$method, drop = TRUE))
simulation_summary <- do.call(rbind, lapply(groups, function(d) {
  data.frame(distribution = d$distribution[1], n = d$n[1], method = d$method[1],
             mean_h = mean(d$h), sd_h = sd(d$h),
             mean_ISE = mean(d$ISE), MCSE = sd(d$ISE) / sqrt(nrow(d)))
}))
rownames(simulation_summary) <- NULL
simulation_summary <- simulation_summary[order(simulation_summary$distribution,
                          simulation_summary$n, simulation_summary$mean_ISE), ]
print(simulation_summary, digits = 3, row.names = FALSE)
cat("CV boundary selections:", sum(simulation_results$CV_boundary &
                                  simulation_results$method == "CV"), "\n")
```

```{r}
#| label: simulation-plots
#| fig-width: 9
#| fig-height: 7
old_par <- par(mfrow = c(2, 2))
for (case in names(examples)) {
  ex <- examples[[case]]
  plot(ex$grid, ex$truth, type = "l", lwd = 3, xlim = c(-4, 4),
       ylim = c(0, max(ex$truth, ex$estimates) * 1.05),
       main = paste(case, "example, n = 500"), xlab = "x", ylab = "Density")
  matlines(ex$grid, ex$estimates[, c("AMISE", "CV", "R_SJ")],
           col = cols[c(1, 2, 5)], lty = 1:3, lwd = 2)
  legend("topright", c("Truth", "AMISE", "CV", "R_SJ"),
         col = c("black", cols[c(1, 2, 5)]), lty = c(1, 1:3), cex = 0.7, bty = "n")
}
for (case in names(simulations)) {
  d <- subset(simulation_results, distribution == case & n == 500)
  boxplot(ISE ~ method, data = d, log = "y", las = 2, col = "grey90",
          main = paste(case, "20 repetitions, n = 500"),
          xlab = "", ylab = "ISE (log scale)")
}
par(old_par)
```

In this experiment at $n=500$, mean ISE is approximately 0.00164 for AMISE
and 0.00241 for CV under the normal model. For the mixture, the corresponding
values are 0.02326 and 0.00349: CV reduces mean error by about 85%. The example
plot explains the difference: the normal-reference bandwidth flattens the two
peaks and fills the valley between them. Under the normal model the bandwidth
standard deviation is about 0.025 for AMISE versus 0.176 for CV. Both methods
have lower mean ISE at $n=500$ than at $n=200$ in these experiments.

The comparison distinguishes **accuracy** (mean ISE) from **stability**
(variation of the chosen bandwidths). A stable normal-reference rule can still
be systematically too smooth for a mixture. CV estimates the error criterion
from the sample and can adapt to multiple peaks, but its selected bandwidth
can fluctuate. The table reports both properties, including the results at the
larger sample size. `R_SJ` is a useful more flexible plug-in baseline; `R_ucv`
uses Gaussian least-squares CV before the scale conversion, so equality with
our exact Epanechnikov CV selector is not expected.

## Performance, alternatives, and bottlenecks

The course's [Lecture 3](https://github.com/math-ku/compstat/blob/main/slides/lecture3.qmd)
compares direct, vectorized, and binned estimators. Our KDE vectorizes the sum
over observations and uses $O(nm)$ work for $m$ grid points, with $O(n+m)$ working
memory. A fully vectorized `outer(grid, x, "-")` alternative needs $O(nm)$ memory;
removing a visible loop does not by itself improve performance. R's `density()`
approximates the calculation by binning and FFT convolution.

The normal-reference selector needs only sample summaries. Our CV implementation
stores $n(n-1)/2$ distances once and reuses them, but still needs $O(n^2)$ memory
and $O(qn^2)$ work for $q$ candidate evaluations. For large samples this is the
main limitation. Blockwise pair summation could reduce memory; binned distances
could reduce runtime at the cost of approximation, as in R's selectors.

The following reproducible benchmark uses a common grid and the same kernel
bandwidth. Repeated calls improve the resolution of the base R timer. The CV
timing includes selection; the density timings measure evaluation with a fixed
bandwidth. These are different tasks and their times should be read accordingly.

```{r}
#| label: benchmark
set.seed(42)
bench_x <- rnorm(2048)
time_ms <- function(f, times) {
  f() # Warm up before measuring.
  median(replicate(3, {
    unname(system.time(for (i in seq_len(times)) f())["elapsed"]) * 1000 / times
  }))
}
benchmark <- do.call(rbind, lapply(c(128, 512, 2048), function(n) {
  x <- bench_x[seq_len(n)]
  grid <- seq(min(x) - 0.5, max(x) + 0.5, length.out = 512)
  data.frame(n = n,
    KDE_ms = time_ms(function() kde(x, 0.5, grid = grid), 10),
    R_density_ms = time_ms(function() density(x, bw = 0.5 / sqrt(5),
      kernel = "epanechnikov", from = min(grid), to = max(grid), n = 512), 100),
    AMISE_ms = time_ms(function() bw_amise(x), 100),
    CV_ms = time_ms(function() bw_cv(x, n_grid = 50), 1))
}))
print(benchmark, digits = 3, row.names = FALSE)
```

A short sampling profile identifies where the selector spends time on this
machine. Temporary profiler output is cleaned up after reading it. Sampling
percentages vary between runs. We report self time to identify where computation
actually occurs, rather than the surrounding Quarto execution calls.

```{r}
#| label: profile-cv
profile_cv <- function(x) {
  path <- tempfile(fileext = ".out")
  on.exit({ Rprof(NULL); unlink(path) }, add = TRUE)
  Rprof(path, interval = 0.005)
  for (i in seq_len(3)) invisible(bw_cv(x, n_grid = 50))
  Rprof(NULL)
  head(summaryRprof(path)$by.self, 8)
}
profile_cv(bench_x[1:1024])
```

The implementation separates validation, kernel evaluation, bandwidth selection,
and diagnostics. This makes it possible to test each mathematical component and
to replace a costly component later. In particular, the numerical CV check
provides a reference for any future binned or blockwise implementation.
