---
title: "Week 1 - CompStat"
author: "Valdemar Schaaf - CLF125"
date: today
format:
  html:
    toc: true
execute:
  echo: true
---

```{r}
#| echo: true
#| message: false
#| warning: false
#| results: hide

library(tidyverse)
```

# Week 1

## Exercise A.1

Explain the result of evaluating the following R expression.

```{r}
(0.1 + 0.1 + 0.1) > 0.3
```

## Exercise A.2

Write a function that takes a numeric vector x and a threshold value h as arguments and returns the vector of all values in x greater than h. Test the function on seq(0, 1, 0.1) with threshold 0.3. Have the example from Exercise A.1 in mind

```{r}
values_greater_than <- function(x, h, tolerance = sqrt(.Machine$double.eps)) {
  valid_values <- !is.na(x) & !is.nan(x) & is.finite(x)
  x[valid_values & x > h + tolerance]
}

values_greater_than(seq(0, 1, 0.1), 0.3)
```

## Exercise A.3 
Investigate how your function from Exercise A.2 treats missing values (NA), infinite values (Inf and -Inf) and the special value “Not a Number” (NaN). Rewrite your function (if necessary) to exclude all or some of such values from x.

```{r}
# Test with missing values
values_greater_than(c(0.1, 0.2, NA, 0.4), 0.25)

# Test with infinite values
values_greater_than(c(0.1, 0.2, Inf, -Inf), 0.25)

# Test with NaN
values_greater_than(c(0.1, 0.2, NaN, 0.4), 0.25)
```

## Data

```{r}
infrared <- read.table("Data/infrared.txt", header = TRUE)
F12 <- infrared$F12
head(F12)
```

## Exercise A.4 
Plot a histogram of log(F12) using the default value of the argument breaks. Experiment with alternative values of breaks.

```{r}
hist(log(F12))
hist(log(F12), breaks = 50)
hist(log(F12), breaks = 100)
```
## Exercise A.5 
Write your own function, called my_breaks, which takes two arguments, x (a vector) and h (a positive integer). Let h have default value 5. The function should first sort x into increasing order and then return the vector that: starts with the smallest entry in x; contains every h'th unique entry from the sorted x; ends with the largest entry in x.

```{r}
my_breaks <- function(x, h = 5) {
  unique_x <- sort(unique(x))
  breaks <- unique_x[seq(1, length(unique_x), by = h)]

  sort(unique(c(
    unique_x[1],
    breaks,
    unique_x[length(unique_x)]
  )))
}
```

```{r}
x = c(1, 3, 2, 5, 10, 11, 1, 1, 3)
my_breaks(x, h = 2)
```

 
Use your function to construct breakpoints for the histogram for different values of h, and compare with the histograms obtained in Exercise A.4.


## Exercise 1: Create a simple s3 class
Write a function make_point(x,y) that returns an object if class "point" with elements x and y.

Write a print method print.point() that prints the point in the form (x,y).

```{r}
make_point <- function(x, y) {
  structure(
    list(x = x, y = y),
    class = "point"
  )
}

print.point <- function(point, ...) {
  cat("(", point$x, ",", point$y, ")\n", sep = "")
}

p <- make_point(2, 3)
p

```
